Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-12-16 src/tex/ms.tex
[showyourwork] overleaf sync

Next, we chose the \textsc{power law + peak} model as the source-frame population model $\Lambda$ and the $\Lambda$CDM model as the cosmological model $\Omega$.
For the redshift population model, instead of the uniform distribution over comoving volume and source-frame time, we used the \textsc{power law} redshift evolution model described in \cite{KAGRA:2021duu}.
Similar to Sec.~\ref{sec:inference_multi}, we fixed some parameters of the population model and the cosmological model.
For the cosmological model, we fixed the parameters to the results from \cite{Planck:2018vyg} except $H_0$.
For the population model, we fixed $\alpha$, $w$, $m_\mathrm{min}$, $m_\mathrm{max}$, and $\delta$ to the results from \cite{LIGOScientific:2021aug}.
Note that the fixed parameters are not necessarily the true values, but they can be considered as a reasonable guess for the parameters.
The fixed parameters of the population model are shown in Table~\ref{tab:fixed_parameters}.
\begin{table}[htbp]
    \caption{Fixed parameters of the $\Lambda$CDM model (left) and the \textsc{power law + peak} model (right).}
    \begin{tabular}{cc|cc}
        \toprule
        Parameter & Value & Parameter & Value  \\
        \midrule
         & & $w$ & $0.024$ \\
        $\Omega_m$ & $0.315$ & $\alpha$ & $4.2\ \mathrm{M}_\odot$\\
        $\Omega_\Lambda$ & $0.685$ & $m_\mathrm{min}$ & $5\ \mathrm{M}_\odot$ \\
        $w$ & $-1$ & $m_\mathrm{max}$ & $109\ \mathrm{M}_\odot$ \\
         & & $\delta$ & $4.9\ \mathrm{M}_\odot$\\
        \botrule
    \end{tabular}
    \label{tab:fixed_parameters}
\end{table}

We then perform the remaining steps similar to Sec.~\ref{sec:inference_multi}.
We set the bounds for the power-law index of the redshift population model $\kappa$ to $[-100, 100]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{KAGRA:2021duu}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}
The result implies that \variable{output/real_H0.txt}, with symmetric $90\%$-credible intervals around the median.
The uncertainty is smaller than the one reported in \cite{KAGRA:2021duu} ($H_0 = 50^{+37}_{-30} \, \mathrm{km/s/Mpc}$), mainly because it only includes the uncertainty from the reconstruction, which underestimates the true uncertainty which also includes the mismatch between the population model and the observed population distribution.
Figure \ref{fig:real_result_multi} shows a peak in the marginalized posterior distribution of $H_0$ around $70\, \mathrm{km/s/Mpc}$.
This means a significant fraction of the reconstructed observed population distribution has an optimized $H_0$ around $70\, \mathrm{km/s/Mpc}$, which again does not include the uncertainty from the mismatch between the population model and the observed population distribution.
The peak should be interpreted as an artifact of the optimization procedure.

% This analysis we report here is a proof of concept of the method presented in this paper, and we simply aim to demonstrate that our method can be applied to real data with relatively little effort.
% To perform a detailed analysis, a more comprehensive configuration that considers different population models, cosmological models, and selection functions is required, as well as a more robust optimization scheme to include more astrophysical parameters.

@@

In this paper, we proposed a method to infer both the intrinsic population model and the cosmological model simultaneously by leveraging the detector-frame mass population distribution and making use of non-parametric methods.
We demonstrated the method with a mock data study and applied it to the real data from the \ac{LVK}.
With the numbers of observed \ac{GW} events that are expected to grow larger and larger, potentially hitting the thousand mark by the end of the fifth observing run, performing a complete analysis from scratch with different astrophysical models will soon become too computationally expensive.
The method we propose in this paper can help mitigate this issue by providing a fast and efficient way to explore different models before performing the full hierarchical Bayesian inference, especially keeping in mind that, if a wrong population model is chosen, the inference result will be biased \citep{Mukherjee:2021rtw, Mastrogiovanni:2021wsd, Pierra:2023deu, LIGOScientific:2020kqk, KAGRA:2021duu}.
This makes our method a powerful diagnostic tool for model selection.

>>>>

Next, we chose the \textsc{power law + peak} model as the source-frame population model $\Lambda$ and the $\Lambda$CDM model as the cosmological model $\Omega$.
For the redshift population model, instead of the uniform distribution over comoving volume and source-frame time, we used the \textsc{power law} redshift evolution model described in Eq.~8 of \cite{KAGRA:2021duu}.
Similar to Sec.~\ref{sec:inference_multi}, we fixed some parameters of the population model and the cosmological model.
For the cosmological model, we fixed the parameters to the results from \cite{Planck:2018vyg} except $H_0$.
For the population model, we fixed $\alpha$, $w$, $m_\mathrm{min}$, $m_\mathrm{max}$, and $\delta$ to the results from \cite{LIGOScientific:2021aug}.
Note that the fixed parameters are not necessarily the true values, but they can be considered as a reasonable guess for the parameters.
The fixed parameters of the population model are shown in Table~\ref{tab:fixed_parameters}.
\begin{table}[htbp]
    \caption{Fixed parameters of the $\Lambda$CDM model (left) and the \textsc{power law + peak} model (right).}
    \begin{tabular}{cc|cc}
        \toprule
        Parameter & Value & Parameter & Value  \\
        \midrule
         & & $w$ & $0.024$ \\
        $\Omega_m$ & $0.315$ & $\alpha$ & $4.2\ \mathrm{M}_\odot$\\
        $\Omega_\Lambda$ & $0.685$ & $m_\mathrm{min}$ & $5\ \mathrm{M}_\odot$ \\
        $w$ & $-1$ & $m_\mathrm{max}$ & $109\ \mathrm{M}_\odot$ \\
         & & $\delta$ & $4.9\ \mathrm{M}_\odot$\\
        \botrule
    \end{tabular}
    \label{tab:fixed_parameters}
\end{table}

We then perform the remaining steps similar to Sec.~\ref{sec:inference_multi}.
We set the bounds for the power-law index of the redshift population model $\kappa$ to $[-100, 100]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{LIGOScientific:2021aug}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}
The result implies that \variable{output/real_H0.txt}, with symmetric $90\%$-credible intervals around the median.
The uncertainty is slightly smaller than the one reported in \cite{LIGOScientific:2021aug} ($H_0 = 50^{+37}_{-30} \, \mathrm{km/s/Mpc}$), mainly because it only includes the uncertainty from the reconstruction, which underestimates the true uncertainty which also includes the mismatch between the population model and the observed population distribution.
Figure \ref{fig:real_result_multi} shows a peak in the marginalized posterior distribution of $H_0$ around $140\, \mathrm{km/s/Mpc}$.
This means a significant fraction of the reconstructed observed population distribution has an optimized $H_0$ around $140\, \mathrm{km/s/Mpc}$, which again does not include the uncertainty from the mismatch between the population model and the observed population distribution.

@@

In this paper, we proposed a method to infer both the intrinsic population model and the cosmological model simultaneously by leveraging the detector-frame mass population distribution and making use of non-parametric methods.
We demonstrated the method with a mock data study and applied it to the real data from the \ac{LVK}.
With the numbers of observed \ac{GW} events that are expected to grow larger and larger, potentially hitting the thousand mark by the end of the fifth observing run, performing a complete analysis from scratch with different astrophysical models will soon become too computationally expensive.
The method we propose in this paper can help mitigate this issue by providing a fast and efficient way to explore different models before performing the full hierarchical Bayesian inference, especially keeping in mind that, if a wrong population model is chosen, the inference result will be biased \citep{Mukherjee:2021rtw, Mastrogiovanni:2021wsd, Pierra:2023deu, LIGOScientific:2020kqk, KAGRA:2021duu, LIGOScientific:2021aug}.
This makes our method a powerful diagnostic tool for model selection.
